---
title: "Groupstage Processing"
author: "Alfie Smith"
date: "28 June 2018"
output: html_document
---

##Overview 

The aim of this script is to convert match-day summaries 
into group-stage results for each team. 

Packages required

* dplyr
* tidyr
* readr
* lubridate
* knitr

Input details: 

* variable name: wc.matches 
* source file: WorldcupMatches.csv 
* fields: home team, away team, home goals, away goals, stage...

Output details:

* variable name: wc.groupstage.totals  
* fields: goals scored, goals conceded, group position...
* output file: wc_groupstage.csv

# Processing

## create Knittr namespace 
For technical reasons to do with R namespaces,
Knittr cannot access the objects in the general R-session. 
We'll need to all load packages and data into memory once again.

```{r message=FALSE, warning=FALSE, cache=FALSE}
# load libs into Knitter's namespace
library(dplyr)
library(tidyr)
library(readr)

# load source data
wc.matches <- read_csv("../data/source_data/WorldCupMatches.csv")

```

## Part 1: clean data and gather into teams 

This processing step renames most of the original columns
to a standardised format (no spaces, lower case)

Filter only the columns that are relevant to this analysis

Convert data from wide format where the rows contain 
both home and away team data to a tidy format 
containing data about only one team.

This last step is critical to rolling up group
stage results per team. 

```{r setup, message=FALSE, warning=FALSE, cache=FALSE}

wc.groupstage.tidy <- wc.matches %>%
  dplyr::rename(year = Year,
         date = Datetime,
         stage = Stage, 
         home = `Home Team Name`,
         away = `Away Team Name`,
         home_goals = `Home Team Goals`,
         away_goals = `Away Team Goals`) %>%
  select(year, date, stage, home, away, home_goals, away_goals) %>%
  
  # convert wide per match data format to skinny per team format. 
  gather("where", "team", c(home, away)) 
```

## Part 2: Calculate new fields 

The source data contains match level summaries that do not 
assign particular values to each team. 

This processing step introduces the concepts of win, lose and draw.

In addition to introducing results, this step
also converts home and away goals to goals scored and conceded. 

Furthermore, I've created a new binary classifier for group stage matches
This is used at the end of the step to remove non-group stage games. 


```{r message=FALSE, warning=FALSE, cache=FALSE}
wc.groupstage.results <- wc.groupstage.tidy %>% 
  # Create new fields: calc per team metrics from match level metrics
  mutate(
        # work out from goals for vs agaist who won the match
        # assign points so we can compare track records. 
        result =  ifelse(  (where == "home" & home_goals > away_goals)|
                            (where == "away" & away_goals > home_goals), 3,
                            ifelse(home_goals==away_goals, 1,
                            0)),
         
         # assign goals per team rather than by match. 
         goals_scored = ifelse(where=="home", home_goals, away_goals),
         goals_conceded = ifelse(where=="home", away_goals, home_goals),
         
         # label whether a match was in group stage or not.
         
         group_stage = ifelse(grepl("Group", stage),TRUE, FALSE)) %>%
  # dataset has many NA rows - remove these. 
  # we're only interested in group stage goals. 
  filter(!is.na(year), 
        group_stage == TRUE) 
```

# Part 3 summarising group results 

This step produces our final product - 

Remove the source and intermediary columns that are not longer required. 
For each team, at each world cup, 

Summarise points and goals scored/conceded for whole group stage. 
Using these totals, rank each teams performance relative to their each group. 

```{r message=FALSE, warning=FALSE, cache=FALSE}

wc.groupstage.totals <- wc.groupstage.results %>%
  # remove intermediary columns 
  select(-home_goals, -away_goals, -group_stage) %>%
  group_by(team, year) %>%
  summarise(points_total = sum(result),
            goals_scored = sum(goals_scored),
            goals_conceded = sum(goals_conceded),
            group_name = first(stage)) %>%
  ungroup() %>%
  group_by(year, group_name) %>%
  # rank teams within group by points total
  # tiny but effective hack -(points total) instead of desc (not working)
  arrange(year, group_name, -(points_total)) %>%
  mutate(rank=row_number()) %>%
  # repeat lines 143-144 for goals scored and conceded
  arrange(year, group_name, -(goals_scored)) %>%
  mutate(goals_scored_rank = row_number()) %>%
  arrange(year, group_name, goals_conceded) %>%
  mutate(goals_conceded_rank = row_number()) 
  
  
```

# Notes from processing 

## Missing data  
We're missing group stage data from the 1934-1938 world cups 
because there were no group stages. These tournaments were 16 teams
only due to some countries refusal to participate in Spainish and Italian WC.

We're missing worldcup data from 1942 and 1946 due to the outbreak
of WW2 causing the cup to be cancelled. 

## Nulls in original data set

There are several thousand null rows in worldcupmatches.csv 
due to an additional comma at the end of the file. 
Instead of removing this manually and created a new source data 
file, I've allowed these rows to be created and filtered them out. 

## > 9 points in some group stages.

The custom for 3 group games has been a standard since 1990 world cup. 
This would lead to a maximum of 9 group stage points.

However, this doesn't hold over the whole time period sampled. 
The 1950 world cup had 3 groups of 6 
Both the 1982 and 1986 world cups operated a second round-robin group-stage 

This inconsistency in maximum points led me to use points ranking 
rather than points total to compare group stage perfomances over time. 




