
---
title: "Placement Processing Notes"
output: html_document
---


## Overview 
Initial dataset has every world cup from 1930-2014 as its entities 

The data is in a wide-format with metrics for each world cup
spread across the row as columns. 

The desired output is a tidy('er') dataset of placements at each worldcup. 
Where each row represents a placement at a given world cup. 
For example 'Germany 2014 Winner'. 

Packages required

* dplyr
* tidyr
* readr
* lubridate
* knitr

Input details: 

* variable name: wc.summary
* source file: WorldCups.csv 
* fields: "Year","Country", "Winner","Runners.Up", "Third" 

Output details:

* variable name: wc.summary.tidy
* fields: "team", "year", "placement", "placement_label", "host"
* output file: wc_placements.csv

# Processsing 



## Preliminary step: create Knittr namespace 
For technical reasons to do with R namespaces,
Knittr cannot use the objects that we've loaded into 
the R session previously. 

We'll need to load all packages again in memory. 

```{r message=FALSE, warning=FALSE, cache=FALSE}
# load libs into Knitter's namespace
wc.summary <- read.csv("../data/source_data/WorldCups.csv")
library(dplyr)
library(tidyr)
```


## Part 1: Clean up of initial wc.summary data 

This section simply renames the columns to my standardised format
(no spaces or caps) and filters out the columns that aren't of
interest to this analysis. 

```{r message=FALSE, warning=FALSE, cache=FALSE}
wc.summary.clean <- wc.summary %>%
  rename(
    year = Year,
    host = Country,
    winner  = Winner, 
    runner_up = Runners.Up,
    third = Third,
    fourth = Fourth,
    teams = QualifiedTeams) %>%
  select(year, host, teams, winner, runner_up, third, fourth) 
  
```


## Part 2: gathering wc.summary.clean into tidy format

This section takes the placement data spread out of several columns, 
(winner, runner up, third, fourth) and gathers it 
into two new columns called 'team' and 'placement label'.

I've also created a numeric version of the placement label variable
where the label 'winner' is converted to a rank '1', which represents 1st. 

```{r message=FALSE, warning=FALSE, cache=FALSE}

wc.summary.tidy <- wc.summary.clean %>%
  # tidy dataset into placements at worldcups
  gather('placement_label','team', 4:7) %>%
  
  # convert placement to a rank integer (1st...4th)
  
  mutate(placement = ifelse(placement_label == "winner", 1,
                     ifelse(placement_label == 'runner_up', 2, 
                     ifelse(placement_label == "third", 3, 4))))
```


# Notes from processing 

## Third awarded to both semi-finalists

The third place play off was introduced in 1970 
world cup. Before this date, both semi-finalists
were treated as 3/4th place finishers. 
The source data writer chose third based on 
goal difference and I've stuck with that rule. 




Script end. 



  
    
